<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Second clutch Bayesian analysis}
-->

Supplementary Materials S2 to 
Karagicheva J., Liebers, M., Rakhimberdiev, E., Hallinger K.K., Saveliev A., A., Winkler, D. W. 2015 Differences in size between first and replacement clutches match the seasonal decline in single clutches in Tree Swallows _Tachycineta bicolor_. - Ibis 000: 000-000.

Depends:  R (>= 3.2.0), JAGS (>=4.0.0)

Authors: Eldar Rakhimberdiev <eldar@nioz.nl> and Julia Karagicheva <jkaraj@gmail.com>

Date: 24.11.2015

License: GNU GPL v3.0

Permanent address: https://github.com/eldarrak/second_clutch/blob/master/S2.rmd

## 1. Get data from Github
```{r, eval = F}
library(RCurl)

text <-getURL("https://raw.githubusercontent.com/eldarrak/second_clutch/master/raw_data.csv",
           ssl.verifypeer = FALSE, followlocation = TRUE)
raw_data<-read.csv(text=text, stringsAsFactors =F)
str(raw_data)
```
In the data frame we have following columns:

1. ```$nseason``` - Year
2. ```$jd``` - julian date
3. ```$datediff``` - difference between second and first clutches
4. ```$nclutch``` - observed clutch size
5. ```$FirstClutchDoubleBreeders``` - 1 if this is the first clutch of a double breeder, otherwise 0.

Note, that ```$datediff``` is 0 for all the first clutches.

## 2. Format data for the analysis
```{r, eval = F}
Res<-data.frame(Year=as.numeric(as.factor(raw_data$nseason)), 
	DayFirst=scale(raw_data$jd, scale=F), DayDiff=raw_data$datediff,
	Clutch=raw_data$nclutch, SingleBreeders=ifelse(raw_data$FirstClutchDoubleBreeders==0 & raw_data$datediff==0, 1, 0),
	FirstClutchDoubleBreeders=raw_data$FirstClutchDoubleBreeders)
```
Here we make variable ```$Year``` that is just a numeric year starting from 1.
We also center ```$jd``` but not scale it
```{r, eval = F}
plot(jitter(log(Res$Clutch))~Res$DayFirst)
```

## 3. Main model w/o variable selection
### Bayesian model
```{r, eval = F}
sink("two_truncated_Poisson_model.jags")
cat("
model {

# model
for (row in 1:nrow) {
log(lamda[row])<- Intercept*SingleBreeders[row]+InterceptFirstClutchDouble*FirstClutchDoubleBreeders[row]+ # Intercept fixed
      InterceptYear[year[row]]*SingleBreeders[row] + #Intercept random
      (Slope+SlopeYear[year[row]])*SingleBreeders[row]*DayFirst[row] + # slope for first clutches
	(Slope+SlopeYear[year[row]]+ SlopeSecondClutch)*DayDiff[row]
}

for (t in 1:nyears) {
InterceptYear[t]~dnorm(0, tau_intercept)
SlopeYear[t]~dnorm(0, tau_slope)
}

#priors
#w_int_double~dbern(0.5)
#w_int~dbern(0.5)
#w_slope~dbern(0.5)
Intercept ~ dunif(-10,10)
InterceptFirstClutchDouble ~ dunif(-10,10)
Slope ~dunif(-0.5,0.5)
SlopeSecondClutch ~dunif(-1,1)

tau_intercept<-pow(sigma_intercept, -2)
sigma_intercept ~ dunif(0,0.5)
tau_slope<-pow(sigma_slope, -2)
sigma_slope ~ dunif(0,1)

# likelihood
for(row in 1:nrow) {
      Clutch[row] ~ dpois(lamda[row]) T(3,)
}
}
",fill = TRUE)
sink()
```

### Inits and data
```{r, eval = F}
nyears=length(unique(Res$Year))

Data <- list("nrow"=nrow(Res), "nyears"=nyears, "year"=Res$Year, "DayFirst"=Res$DayFirst, 
        "DayDiff"=Res$DayDiff, "Clutch"=Res$Clutch,
		"SingleBreeders"=SingleBreeders,
        "FirstClutchDoubleBreeders"=Res$FirstClutchDoubleBreeders)

inits <- function() list("Intercept" = runif(1,-1,3),  
                         "InterceptFirstClutchDouble"= runif(1,-1,3),
                         "InterceptYear" = runif(nyears,-1,1),
                         "Slope" =-runif(1,-0.5, 0.5),
                         "SlopeYear" =runif(nyears,-1,1),
                         "SlopeSecondClutch" =runif(1,-1,1),
                         "sigma_intercept"=runif(1, 0,0.5),
                         "sigma_slope"=runif(1,0,1))

# Parameters monitored
params <- c("Intercept", "InterceptFirstClutchDouble", "InterceptYear",
            "Slope", "SlopeYear", "SlopeSecondClutch", "sigma_intercept", "sigma_slope")
```

### Model run
This will take consideravle amount of time (a few hours), lower ```n.iter``` if you want it to finish faster!
```{r, eval = F}
library(R2jags)
# Call JAGS from R (few hours)

out <- jags.parallel(Data, inits, params, "two_truncated_Poisson_model.jags",
                     n.chains = 5, n.thin = 500, n.iter = 250000, n.burnin = 50000)

save(out, file="model_output.RData")
```
### Exploring results

```{r, eval = F}
print(out)

# ok, now overall slope and intercept
Intercept_pred<-quantile(out$BUGSoutput$sims.list$Intercept, c(0.025, 0.5, 0.075)) 
Slope_pred<-quantile(out$BUGSoutput$sims.list$Slope, c(0.025, 0.5, 0.075)) 

XX<-seq(min(Res$DayFirst)-10, max(Res$DayFirst)+10)
Y_main<-Intercept_pred[2]+Slope_pred[2]*XX
# now CI

Y_lower<-apply(out$BUGSoutput$sims.list$Slope %*% XX + 
               out$BUGSoutput$sims.list$Intercept[,1],
               2, quantile, c(0.025))

Y_upper<-apply(out$BUGSoutput$sims.list$Slope %*% XX +
               out$BUGSoutput$sims.list$Intercept[,1],
               2, quantile, c(0.975))

Y_second_clutch<-apply(out$BUGSoutput$sims.list$Slope %*% XX +
                       out$BUGSoutput$sims.list$Intercept[,1]+
                       out$BUGSoutput$sims.list$Intercept_first_clutch_double[,1]+
                       out$BUGSoutput$sims.list$SlopeSecondClutch %*% XX,
                       2, quantile, c(0.5))

Y_second_clutch_upper<-apply(out$BUGSoutput$sims.list$Slope %*% XX +
                             out$BUGSoutput$sims.list$Intercept[,1]+
                             out$BUGSoutput$sims.list$Intercept_first_clutch_double[,1]+
                             out$BUGSoutput$sims.list$SlopeSecondClutch %*% XX,
                             2, quantile, c(0.975))
Y_second_clutch_lower<-apply(out$BUGSoutput$sims.list$Slope %*% XX +
                             out$BUGSoutput$sims.list$Intercept[,1]+
                             out$BUGSoutput$sims.list$Intercept_first_clutch_double[,1]+
                             out$BUGSoutput$sims.list$SlopeSecondClutch %*% XX,
                             2, quantile, c(0.025))

pdf("figure1.pdf")

plot(jitter(Res$Clutch)~jitter(Res$DayFirst), pch=".", cex=2,
      xlab="Time (days)", ylab="Clutch size", las=1, type="n", xlim=c(Res$Clutch))

points(jitter(Res$Clutch)~jitter(Res$DayFirst), pch=".", cex=2, col=grey(0.2))


polygon(c(XX, rev(XX)), exp(c(Y_upper, rev(Y_lower)))
	, col=grey(0.4, 0.5), border=grey(0.1), lwd=2)
#lines(exp(Y_upper)~XX)
#lines(exp(Y_lower)~XX)
polygon(c(XX, rev(XX)), exp(c(Y_second_clutch_upper, rev(Y_second_clutch_lower))), 
	col=grey(0.6, alpha=0.5), border=grey(0.99), lwd=2)

lines(exp(Y_main)~XX, lwd=3, col=grey(0.1))

lines(exp(Y_second_clutch)~XX, col=grey(0.99), lwd=3, lty=1)


box()

dev.off()

#lines(exp(Y_second_clutch_lower)~XX, lty=2)
#lines(exp(Y_second_clutch_upper)~XX, lty=2)
```

## 4. Model with variable selection
For the variables `Intercept_first_clutch_double`, `InterceptYear`, `SlopeYear`) introduce additional parameters that can take values only of 1 or 0.
### Bayesian model
```{r, eval = F}
sink("two_truncated_Poisson_model_with_MS.jags")
cat("
model {

# model
for (row in 1:nrow) {
log(lamda[row])<- Intercept*SingleBreeders[row]+w_int_double*InterceptFirstClutchDouble*FirstClutchDoubleBreeders[row]+
                  w_int*InterceptYear[year[row]]*SingleBreeders[row]+ #intercept
                  (Slope+w_slope*SlopeYear[year[row]])*SingleBreeders[row]*DayFirst[row] + # slope for first clutches
	          (Slope+w_slope*SlopeYear[year[row]]+ w_slope_second*SlopeSecondClutch)*DayDiff[row]
}

for (t in 1:nyears) {
InterceptYear[t]~dnorm(0, tau_intercept)
SlopeYear[t]~dnorm(0, tau_slope)
}

#priors
w_int_double~dbern(0.5)
w_int~dbern(0.5)
w_slope~dbern(0.5)
w_slope_second~dbern(0.5)
Intercept ~ dunif(-10,10)
InterceptFirstClutchDouble ~ dunif(-10,10)
Slope ~dunif(-0.5,0.5)
SlopeSecondClutch ~dunif(-1,1)

tau_intercept<-pow(sigma_intercept, -2)
sigma_intercept ~ dunif(0,0.5)
tau_slope<-pow(sigma_slope, -2)
sigma_slope ~ dunif(0,1)

# likelihood
for(row in 1:nrow) {
      Clutch[row] ~ dpois(lamda[row]) T(3,)
}
}
",fill = TRUE)
sink()
```

###Inits and data
```{r, eval = F}
nyears=length(unique(Res$Year))

Data <- list("nrow"=nrow(Res), "nyears"=nyears, "year"=Res$Year, "DayFirst"=Res$DayFirst, 
        "DayDiff"=Res$DayDiff, "Clutch"=Res$Clutch,
		"SingleBreeders"=SingleBreeders,
        "FirstClutchDoubleBreeders"=Res$FirstClutchDoubleBreeders)
		
inits <- function() list("Intercept" = runif(1,-1,3),
                         "InterceptFirstClutchDouble"= runif(1,-1,3),
                         "InterceptYear" = runif(nyears,-1,1),
                         "Slope" =-runif(1,-0.5, 0.5),
                         "SlopeYear" =runif(nyears,-1,1),
                         "SlopeSecondClutch" =runif(1,-1,1),
                         "sigma_intercept"=runif(1, 0,0.5),
                         "sigma_slope"=runif(1,0,1),
                         "w_int"=rbinom(1, 1, 0.5),
                         "w_int_double"=rbinom(1, 1, 0.5),
                         "w_slope"=rbinom(1, 1, 0.5),
                         "w_slope_second"=rbinom(1, 1, 0.5))

# Parameters monitored
params <- c("Intercept", "InterceptFirstClutchDouble", "InterceptYear",
            "Slope", "SlopeYear", "SlopeSecondClutch",
            "sigma_intercept", "sigma_slope",
            "w_int", "w_int_double","w_slope",
			"w_slope_second")
```

### Model run
This will take consideravle amount of time (~0.5 hour), lower ```n.iter``` if you want it to finish faster!
```{r, eval = F}
library(R2jags)
# Call JAGS from R (~0.5 hour)
out_ms <- jags.parallel(Data, inits, params, "two_truncated_Poisson_model_with_MS.jags",
                        n.chains = 5, n.thin = 500, n.iter = 250000, n.burnin = 50000)
save(out_ms, file="model_selection_output.RData")
```
