<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Second clutch Bayesian analysis}
-->

Supplementary Materials S2 to 
Karagicheva J., Liebers, M., Rakhimberdiev, E., Hallinger K.K., Saveliev A., A., Winkler, D. W. 2015 Differences in size between first and replacement clutches match the seasonal decline in single clutches in Tree Swallows **Tachycineta bicolor**. - Ibis 000: 000-000.

Depends:  
	R (>= 3.2.0),
	JAGS (>=4.0.0)
Authors: Eldar Rakhimberdiev <eldar@nioz.nl> and Julia Karagicheva <jkaraj@gmail.com>
License: GNU GPL v3.0

# 1. Get data from Github
```{r}
library(RCurl)

text <-getURL("https://raw.githubusercontent.com/eldarrak/second_clutch/master/raw_data.csv",
           ssl.verifypeer = FALSE, followlocation = TRUE)
raw_data<-read.csv(text=text, stringsAsFactors =F)
str(raw_data)
```
In the data frame we have following columns:
1. ```$nseason``` - Year
2. ```$jd``` - julian date
3. ```$datediff``` - difference between second and first clutches
4. ```$nclutch``` - observed clutch size
Note, that ```$datediff``` is 0 for all the first clutches.

# 2. Format data for the analysis
```{r}
Res<-data.frame(Year=as.numeric(as.factor(allbreeders$nseason)), 
	DayFirst=scale(allbreeders$jd, scale=F), DayDiff=allbreeders$datediff,
	Clutch=allbreeders$nclutch)
```
Here we make variable ```$Year``` that is just a numeric year starting from 1.
We also center ```$jd``` but not scale it

plot(jitter(log(Res$Clutch))~Res$DayFirst)


# 3. Bayesian model
```{r}
sink("two_truncated_Poisson_model_with_MS.jags")
cat("
model {

# model
for (row in 1:nrow) {
log(lamda[row])<- Intercept+w_int*InterceptYear[year[row]]+ #intercept
    (Slope+w_slope*SlopeYear[year[row]])*Day[row] + # slope for first clutches
	(Slope+w_slope*SlopeYear[year[row]]+ SlopeSecondClutch)*DayDiff[row]
}

for (t in 1:nyears) {
InterceptYear[t]~dnorm(0, tau_intercept)
SlopeYear[t]~dnorm(0, tau2)
}

#priors
w_int~dbern(0.5)
w_slope~dbern(0.5)
Intercept ~ dunif(-10,10)
Slope ~dunif(-0.5,0.5)
SlopeSecondClutch ~dunif(-1,1)

tau_intercept<-pow(sigma_intercept, -2)
sigma_intercept ~ dunif(0,0.5)
tau_slope<-pow(sigma_slope, -2)
sigma_slope ~ dunif(0,1)

# likelihood
for(row in 1:nrow) {
      Clutch[row] ~ dpois(lamda[row]) T(2,)
}
}
",fill = TRUE)
sink()
```

# 3. Preparing inits and data
```{r}

nyears=length(unique(Res$Year))

Data <- list(nrow=nrow(Res), nyears=nyears, year=Res$Year, DayFirst=Res$DayFirst, DayDiff=Res$DayDiff, Clutch=Res$Clutch)

inits <- function() list("Intercept" = runif(1,-1,3),  "InterceptYear" = runif(nyears,-1,1), "Slope" =-runif(1,-0.5, 0.5),"SlopeYear" =runif(nyears,-1,1), "SlopeSecondClutch" =runif(1,-1,1), "sigma_intercept"=runif(1, 0,0.5), "sigma_slope"=runif(1,0,1), w_int=rbinom(1, 1, 0.5), w_slope=rbinom(1, 1, 0.5))

# Parameters monitored
params <- c("Intercept", "InterceptYear", "Slope", "SlopeYear", "SlopeSecondClutch", "sigma_intercept", "sigma_slope", "w_int", "w_slope")


# 4. Model run
This will take consideravle amount of time (a few hours), lower ```n.iter``` if you want it to finish faster!
```{r}
library(R2jags)
# Call JAGS from R (few hours)

out_ms <- jags.parallel(Data, inits, params, "two_truncated_Poisson_model_with_MS.jags", n.chains = 5, n.thin = 500, n.iter = 250000, n.burnin = 50000, working.directory = getwd())
```

#5. Exploring results

```{r}
out_ms
```









